---
title: "Air Pollution and the Serum Metabolome in Patients with Parkinson's Disease"
author: "Dayoon Kwon"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(readr)
library(readxl)
library(limma)
library(preprocessCore)
library(KEGGREST)
```

# Introduction

This file documents code used to perform the main statistical analyses detailed in the manuscript ***Air Pollution and the Serum Metabolome in Patients with Parkinson's Disease***. The analyses include a metabolome-wide association study (MWAS) of long-term ambient air pollution exposures (PM~2.5~, CO, and NO~2~) with untargeted serum LC-HRMS metabolomic features in PD patients and population-based controls, followed by metabolite annotation and pathway enrichment analysis.

# Setting up R environment for statistical analyses

## Create needed functions for analyses

```{r}
# LIMMA wrapper: build design matrix dynamically and run eBayes
run_limma <- function(metabolic, covariates, covariate_names) {
  formula_str <- paste(" ~ ", paste(covariate_names, collapse = "+"))
  design <- model.matrix(as.formula(formula_str), data = covariates)
  fit <- lmFit(metabolic, design)
  fit <- eBayes(fit)
  return(fit)
}

# Core MWAS function: loops over exposures and fits LIMMA in the full sample
# and within each PEG wave (PEG1, PEG2). Returns log2 fold-changes, CIs, and
# a flag indicating consistent direction with p < 0.05 across PEG1 and PEG2.
run_analysis <- function(exposures, data, matrix_link, matrix_data,
                         covariates, PD_status) {
  all_results <- list()

  for (exposure in exposures) {
    # Build phenotype data for this exposure
    phenotype_data <- data %>%
      left_join(matrix_link %>%
                  rename(PEGID = pegid) %>%
                  filter(type == "baseline") %>%
                  select(PEGID, FullRunName),
                by = "PEGID") %>%
      filter(PD == PD_status) %>%
      select(FullRunName, peg, PD, !!sym(exposure), all_of(covariates)) %>%
      arrange(FullRunName) %>%
      na.omit()
    rownames(phenotype_data) <- phenotype_data$FullRunName

    # Align intensity matrix to phenotype rows
    matrix_prepared <- matrix_data %>%
      t() %>% data.frame() %>%
      filter(rownames(.) %in% phenotype_data$FullRunName) %>%
      arrange(rownames(.)) %>%
      as.matrix() %>% t()

    run_limma_subset <- function(subset_data, covariate_vec) {
      fit <- run_limma(matrix_prepared[, subset_data$FullRunName],
                      subset_data, covariate_vec)
      topTable(fit, coef = exposure, n = Inf, adjust = "fdr", confint = TRUE) %>%
        mutate(
          mz   = sub("mz_rt_([0-9.]+)_([0-9.]+)", "\\1", rownames(.)),
          time = sub(".*_([0-9.]+)$", "\\1", rownames(.)),
          log2FC = logFC / log(2),
          LL = CI.L / log(2),
          UL = CI.R / log(2)
        ) %>%
        select(mz:time, log2FC, LL, UL, AveExpr, t, P.Value, adj.P.Val) %>%
        arrange(adj.P.Val, P.Value)
    }

    # Full sample (include PEG wave as a covariate)
    result_all  <- run_limma_subset(phenotype_data, c(covariates, exposure))

    # PEG1 and PEG2 separately (drop PEG covariate, rename outputs)
    result_peg1 <- run_limma_subset(filter(phenotype_data, peg == 1),
                                    c(setdiff(covariates, "peg"), exposure)) %>%
      rename_with(~ paste0(., "_peg1"), -c(mz, time))
    result_peg2 <- run_limma_subset(filter(phenotype_data, peg == 2),
                                    c(setdiff(covariates, "peg"), exposure)) %>%
      rename_with(~ paste0(., "_peg2"), -c(mz, time))

    final_result <- result_all %>%
      left_join(result_peg1, by = c("mz", "time")) %>%
      left_join(result_peg2, by = c("mz", "time")) %>%
      mutate(
        pvalue_and_direction = ifelse(
          P.Value < 0.05 & sign(log2FC_peg1) == sign(log2FC_peg2), 1, 0),
        PD = PD_status
      )

    all_results[[exposure]] <- final_result
  }

  return(all_results)
}

# Build mz/rt input table for xMSannotator from a feature intensity matrix
annotation_input <- function(data) {
  split_names <- strsplit(rownames(data), "_")
  mz_rt_df <- do.call(rbind, lapply(split_names, function(x) {
    data.frame(mz = as.numeric(x[3]), time = as.numeric(x[4]))
  }))
  cbind(mz_rt_df, data)
}

# Convert ppm tolerance to an absolute mz window
ppm_to_mz <- function(mz, ppm) mz * (ppm / 1e6)

# Parse pollutant / averaging window from MWAS result list element names
extract_info <- function(name, mode) {
  parts <- strsplit(name, "_")[[1]]
  pollutant <- parts[1]; year <- parts[2]
  model <- dplyr::case_when(
    pollutant == "pm"   ~ "ctm",
    pollutant == "co"   ~ "caline4",
    pollutant == "pm25" ~ "lur",
    pollutant == "no"   ~ "lur"
  )
  pollutant_name <- dplyr::case_when(
    pollutant == "pm"   ~ "pm2.5",
    pollutant == "pm25" ~ "pm2.5",
    pollutant == "co"   ~ "co",
    pollutant == "no"   ~ "no2"
  )
  data.frame(mode = mode, pollutant = pollutant_name,
             model = model, year = gsub("y", "", year))
}

# Flatten a per-exposure list of MWAS result tables into one data frame,
# tagging each row with mode/pollutant/model/year via extract_info()
combine_results <- function(results_list, mode_name) {
  bind_rows(lapply(names(results_list), function(name) {
    df <- results_list[[name]]
    meta <- extract_info(name, mode_name)
    cbind(meta, df)
  }))
}
```

# Air pollution exposure data

## Correlation across exposure models and averaging windows

```{r, warning=FALSE, message=FALSE}
# ap_data: subject-level air pollution exposures across 4 models
# (PM2.5 CTM, CO CALINE4, PM2.5 LUR, NO2 LUR) x 3 averaging windows (1Y, 3Y, 5Y).
exposures <- c("pm_1y",  "co_1y",  "pm25_1y", "no_1y",
               "pm_3y",  "co_3y",  "pm25_3y", "no_3y",
               "pm_5y",  "co_5y",  "pm25_5y", "no_5y")

correlation_matrix <- cor(ap_data %>% select(all_of(exposures)) %>% na.omit(),
                          method = "spearman")
```

## Descriptive statistics

```{r, warning=FALSE, message=FALSE}
descriptive_stats <- lapply(exposures, function(var) {
  ap_data %>%
    summarise(
      Variable = var,
      Mean     = mean(.data[[var]], na.rm = TRUE),
      SD       = sd(.data[[var]], na.rm = TRUE),
      Minimum  = min(.data[[var]], na.rm = TRUE),
      Q1       = quantile(.data[[var]], 0.25, na.rm = TRUE),
      Median   = median(.data[[var]], na.rm = TRUE),
      Q3       = quantile(.data[[var]], 0.75, na.rm = TRUE),
      Maximum  = max(.data[[var]], na.rm = TRUE),
      IQR      = IQR(.data[[var]], na.rm = TRUE)
    )
}) %>% bind_rows()
```

## IQR scaling of exposures

```{r, warning=FALSE, message=FALSE}
# Scale each exposure by its interquartile range so effect estimates
# are reported per one-IQR increment.
ap_data2 <- ap_data %>%
  mutate(across(all_of(exposures), ~ ./IQR(., na.rm = TRUE),
                .names = "{.col}_iqr"))

exposures <- ap_data2 %>% select(ends_with("_iqr")) %>% names()
```

# Metabolome-Wide Association Study (MWAS)

## Run LIMMA across all exposures, separately by PD status

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Covariates: age, sex, race/ethnicity, education (years), sample year, PEG wave
covariate_names <- c("agenew", "sex", "minority", "a1_schyrs",
                     "sample_year", "peg")

# pheno: merged phenotype/covariate table including PD status and exposures
# link_hilic / link_c18: maps subject IDs (PEGID) to LC-MS FullRunName for
#   the HILIC-positive and C18-negative modes
# metab_hilic / metab_c18: post-QC, batch-corrected feature intensity matrices
#   (rows = features named "mz_rt_<mz>_<rt>", columns = FullRunName)

# PD cases
results_hilic_PD1 <- run_analysis(exposures, pheno, link_hilic, metab_hilic,
                                  covariate_names, PD_status = 1)
results_c18_PD1   <- run_analysis(exposures, pheno, link_c18,   metab_c18,
                                  covariate_names, PD_status = 1)

# Population-based controls
results_hilic_PD0 <- run_analysis(exposures, pheno, link_hilic, metab_hilic,
                                  covariate_names, PD_status = 0)
results_c18_PD0   <- run_analysis(exposures, pheno, link_c18,   metab_c18,
                                  covariate_names, PD_status = 0)
```

## Combine into a single MWAS results table

```{r, warning=FALSE, message=FALSE, eval=FALSE}
df_PD1 <- bind_rows(combine_results(results_hilic_PD1, "hilic"),
                    combine_results(results_c18_PD1,   "c18"))
df_PD0 <- bind_rows(combine_results(results_hilic_PD0, "hilic"),
                    combine_results(results_c18_PD0,   "c18"))

df <- bind_rows(df_PD1, df_PD0) %>%
  mutate(
    year      = as.integer(year),
    mode      = factor(mode,      levels = c("hilic", "c18")),
    pollutant = factor(pollutant, levels = c("pm2.5", "co", "no2")),
    model     = factor(model,     levels = c("ctm", "caline4", "lur"))
  ) %>%
  arrange(PD, mode, pollutant, model, year, adj.P.Val, P.Value) %>%
  select(mode, pollutant, model, year, PD, everything())

# Flag features that are nominally significant with consistent direction
# in BOTH PD cases and controls.
df2 <- df %>%
  group_by(pollutant, model, year, mode, mz, time) %>%
  mutate(
    shared_signal_flag = ifelse(
      sum(PD == 1 & pvalue_and_direction == 1) > 0 &
      sum(PD == 0 & pvalue_and_direction == 1) > 0, 1, 0
    )
  ) %>%
  ungroup()
```

# PD vs Control comparison

For each feature, compare nominal significance and direction across PD cases and controls. "Both" features (p < 0.05 with consistent direction in both groups) are the target set for shared-response interpretation.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
df_pd <- df2 %>% filter(PD == 1) %>%
  select(mz, time, mode, pollutant, model, year,
         log2FC_PD = log2FC, pval_dir_PD = pvalue_and_direction)
df_ctrl <- df2 %>% filter(PD == 0) %>%
  select(mz, time, mode, pollutant, model, year,
         log2FC_CTRL = log2FC, pval_dir_CTRL = pvalue_and_direction)

df_pdctrl <- left_join(df_pd, df_ctrl,
                       by = c("mz", "time", "mode", "pollutant", "model", "year")) %>%
  mutate(
    sig = case_when(
      pval_dir_PD == 1 & pval_dir_CTRL == 1 ~ "Both",
      pval_dir_PD == 1                       ~ "PD only",
      pval_dir_CTRL == 1                     ~ "Control only",
      TRUE                                   ~ "NS"
    ),
    sig = factor(sig, levels = c("Both", "PD only", "Control only", "NS"))
  )

# Pearson correlation between PD and Control log2FC among "Both" features,
# stratified by pollutant x model
cor_pd_vs_ctrl <- df_pdctrl %>%
  filter(sig == "Both") %>%
  group_by(pollutant, model, year) %>%
  summarise(
    n           = n(),
    correlation = cor(log2FC_PD, log2FC_CTRL, method = "pearson"),
    p_value     = cor.test(log2FC_PD, log2FC_CTRL, method = "pearson")$p.value,
    .groups = "drop"
  )
```

# Replication: PEG1 vs PEG2

```{r warning=FALSE, message=FALSE, eval=FALSE}
df_peg <- df2 %>%
  filter(PD == 1) %>%
  mutate(
    sig = case_when(
      adj.P.Val < 0.05 & sign(log2FC_peg1) == sign(log2FC_peg2) ~ "FDR<0.05, same dir",
      P.Value  < 0.05 & sign(log2FC_peg1) == sign(log2FC_peg2)  ~ "p<0.05, same dir",
      (adj.P.Val < 0.05 | P.Value < 0.05) &
        sign(log2FC_peg1) != sign(log2FC_peg2)                   ~ "p<0.05, opp dir",
      TRUE                                                        ~ "NS"
    )
  )

# Correlation between PEG1 and PEG2 effect sizes among same-direction,
# nominally significant features, stratified by pollutant x model
cor_peg1_vs_peg2 <- df_peg %>%
  filter(sig == "p<0.05, same dir") %>%
  group_by(pollutant, model, year) %>%
  summarise(
    n           = n(),
    correlation = cor(log2FC_peg1, log2FC_peg2, method = "pearson"),
    p_value     = cor.test(log2FC_peg1, log2FC_peg2, method = "pearson")$p.value,
    .groups = "drop"
  )
```

# Pathway enrichment analysis (Mummichog)

## Combine Mummichog output across exposures

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# mummichog_data: list of pathway-enrichment tables, one per exposure folder.
# Each folder name encodes pollutant and averaging window.
mummichog_data2 <- list()
for (name in names(mummichog_data)) {
  pollutant <- strsplit(name, "_")[[1]][1]
  avgyr     <- strsplit(name, "_")[[1]][2]
  mummichog_data2[[name]] <- mummichog_data[[name]] %>%
    mutate(pollutant = pollutant, avgyr = avgyr)
}

mummichog_data3 <- bind_rows(mummichog_data2) %>%
  mutate(EnrichmentFactor = Hits.sig / Expected)

# A pathway is treated as significant when BOTH Hits.sig >= 5 AND
# P.Fisher. < 0.05.
sig_pathways <- mummichog_data3 %>%
  filter(P.Fisher. < 0.05 & Hits.sig >= 5)
```

# Metabolite annotation

## In-house library (tier-1 confidence)

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Filter MWAS results to nominally significant features and build mz/rt
# match windows for library lookup.
mz_tolerance_ppm <- 10   # ppm
time_tolerance_s <- 30   # seconds

annotation_base <- df2 %>%
  filter(P.Value < 0.05) %>%
  select(mode:UL, P.Value, adj.P.Val) %>%
  mutate(mz   = round(as.numeric(mz),   3),
         time = round(as.numeric(time), 3),
         lower_mz   = mz - ppm_to_mz(mz, mz_tolerance_ppm),
         upper_mz   = mz + ppm_to_mz(mz, mz_tolerance_ppm),
         lower_time = time - time_tolerance_s,
         upper_time = time + time_tolerance_s) %>%
  select(mode:time, lower_mz:upper_time, everything())

# In-house HILIC (positive mode) library
inhouse_hilic2 <- annotation_base %>%
  filter(mode == "hilic") %>%
  left_join(inhouse_hilic, by = character()) %>%
  filter(mz_library   >= lower_mz   & mz_library   <= upper_mz,
         time_library >= lower_time & time_library <= upper_time) %>%
  mutate(mz_diff = abs(mz - mz_library),
         rt_diff = abs(time - time_library))

# In-house C18 (negative mode) library — same join pattern
inhouse_c182 <- annotation_base %>%
  filter(mode == "c18") %>%
  left_join(inhouse_c18, by = character()) %>%
  filter(mz_library   >= lower_mz   & mz_library   <= upper_mz,
         time_library >= lower_time & time_library <= upper_time) %>%
  mutate(mz_diff = abs(mz - mz_library),
         rt_diff = abs(time - time_library))

inhouse <- bind_rows(inhouse_hilic2, inhouse_c182)
annotation_base2 <- merge(annotation_base, inhouse) %>% distinct()
```

## Mummichog matched compounds + KEGG names

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# compound_data: list of "mummichog_matched_compound_all" tables per exposure
compound_data2 <- list()
for (name in names(compound_data)) {
  pollutant_info <- strsplit(name, "_")[[1]]
  compound_data2[[name]] <- compound_data[[name]] %>%
    mutate(pollutant = pollutant_info[1], year = pollutant_info[2])
}

compound_data3 <- bind_rows(compound_data2) %>%
  rename(mz = Query.Mass, time = Retention.Time,
         Best_Guess_Compound = Matched.Compound,
         ion = Matched.Form, mz_Diff = Mass.Diff) %>%
  select(pollutant, year, mz, time, Best_Guess_Compound, ion, mz_Diff)

# Look up compound names from KEGG
compound_ids  <- unique(compound_data3$Best_Guess_Compound)
compound_info <- list()
for (compound_id in compound_ids) {
  tryCatch({
    kegg_data <- keggGet(compound_id)
    if (!is.null(kegg_data[[1]])) {
      compound_info[[compound_id]] <- kegg_data[[1]]$NAME[1]
    }
  }, error = function(e) message("Could not retrieve ", compound_id))
}

compound_data4 <- data.frame(
  Best_Guess_Compound = names(compound_info),
  Best_Guess_Name     = unlist(compound_info),
  stringsAsFactors    = FALSE
)
```

## xMSannotator (tier-2/3 confidence)

```{r, eval=FALSE, warning=FALSE, message=FALSE}
library(xMSannotator)
data(adduct_table); data(adduct_weights)

# HILIC positive
dataA <- annotation_input(metab_hilic) %>% unique()
max.mz.diff        <- 10
max.rt.diff        <- 30
corthresh          <- 0.7
max_isp            <- 5
mass_defect_window <- 0.01

multilevelannotation(
  dataA = dataA, max.mz.diff = max.mz.diff, max.rt.diff = max.rt.diff,
  cormethod = "pearson", num_nodes = 2,
  queryadductlist = c("all"), mode = "pos",
  outloc = "xmsAnnotator/hilic",
  db_name = "HMDB", adduct_weights = adduct_weights,
  num_sets = 300, allsteps = TRUE,
  corthresh = corthresh, NOPS_check = TRUE,
  deepsplit = 2, networktype = "unsigned",
  minclustsize = 10, module.merge.dissimilarity = 0.2,
  filter.by = c("M+H"), max_isp = max_isp,
  HMDBselect = "union", mass_defect_window = mass_defect_window,
  pathwaycheckmode = "pm", mass_defect_mode = "pos"
)

# C18 negative — same call with mode = "neg", filter.by = c("M-H"),
# and outloc = "xmsAnnotator/c18"
```

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Stage 5 = consensus annotations. Keep Confidence >= 2.
xms_hilic <- xms_hilic_raw %>% mutate(mode = "positive") %>% filter(Confidence >= 2)
xms_c18   <- xms_c18_raw   %>% mutate(mode = "negative") %>% filter(Confidence >= 2)
xms_data  <- bind_rows(xms_hilic, xms_c18)

xms_matched <- left_join(annotation_base, xms_data, by = "mode") %>%
  filter(lower_mz   <= mz.y   & mz.y   <= upper_mz,
         lower_time <= time.y & time.y <= upper_time) %>%
  select(-c(mz.y, time.y)) %>%
  rename(mz = mz.x, time = time.x)
```
